---
title: "Class 12: HW Population Analysis Pt.2"
author: "Ryan Ellis (PID: A17673864)"
format: pdf
toc: true
---

## Section 1. Proportion of G/G in a population

Downloaded a CSV file from ensemble-->https://www.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#_

```{r}
MXL<- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(MXL)
```

```{r}
round(table(MXL$Genotype..forward.strand.)/nrow(MXL)*100,2)
```

Lets look at the GBR dataset

```{r}
GBR<-read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

```{r}
round(table(GBR$Genotype..forward.strand.)/nrow(GBR)*100,2)
```

This variant (GG) that is associated with childhood asthma is more frequent in GBR populations than MXL populations. 

Further dig. 

## Homework Problems

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
pop<- read.table("rs8067378_ENSG00000172057.6.txt")

nrow(pop)
table(pop$geno)

?median()

```
```{r}
head(pop)
```

```{r}
library(dplyr)
pop|>
  arrange(geno)|>
  group_by(geno)|>
  summarize(MedianExp<-median(exp))
```

Sample size= 462
AA: 108 (median expression:31.2), AG:233(median expression:25.1), GG:121(median expression: 20.1)

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)
ggplot(pop)+
  aes(x=geno, y=exp, fill=geno)+
  geom_boxplot(notch=TRUE)
```
You can infer using the above box plot, that genotyping has a correlation with expression levels. With GG showing lower expression levels than AA. Yes these SNP variation do indeed seem to play a role in ORMDL3 expression, so depending on the variation this will either have an increase or decrease in expression levels.  
