---
title: "Class 09: Candy Mini-Project"
author: "Ryan Ellis (PID: A17673864)"
format: pdf
toc: TRUE
---

## Background

In this mini project we will take a dive into some exploratory techniques within `r`, including creating correlation plots and PCA plots to explore and intrepret a dataset on candy. 

```{r}
candydata<- "candy-data.csv"
candy = read.csv(candydata, row.names=1)
head(candy)
```


> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
table(candy$fruity)
```
There are 38 different fruity types of candy. 

## What is your favorite candy?

The percentage of people who prefer a certain candy over a randomly chosen one is called `$winpercent`.

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["Milky Way", ]$winpercent
```
The winpercent value for Milky Way is 73%

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
The winpercent for Kit Kat is around 77% 

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
The winpercent for Tootsie Roll Snack Bars is around 50%

Now we will dive into the `skim()` function that is apart of the **skimr** package. This helps to summarize and overview the dataset. 

```{r}
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The `winpercent` column is on an entriely different scale than the other columns, while the other coulmns fall between 0-1, the `winpercent` coulumn falls between 0-100. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

The presence or absence of chocolate within that particular candy. A 1 designating that candy has chocolate and a 0 designating that it is chocolate free.

## Exploratory Analysis

> Q8. Plot a histogram of winpercent values using both base R and ggplot2.

```{r}
library(ggplot2)
ggplot(candy)+
  aes(winpercent)+
  geom_histogram(bins=20)
```

> Q9. Is the distribution of winpercent values symmetrical?

No they are not symmetrical, it appears that they are skewed to the right. 

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```
The mean is slightly above 50, while the median is lower than 50. Median is a better tell of the true center of the data. 

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.win<- candy[as.logical(candy$chocolate==1),"winpercent"]
fruit.win<- candy[as.logical(candy$fruity==1),"winpercent"]
mean(choc.win)
mean(fruit.win)

mean(choc.win)>mean(fruit.win)
```

On average choclate candy is ranked higher than fruit candy.

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

p-value = 2.871e-08, so they are significantly different because they are below the alpha of 0.05.

## Overall Candy Rankings 

> Q13. What are the five least liked candy types in this set?

```{r,message=FALSE}
# Here we are using dpylr 
library(dplyr)
candy |>
  arrange(winpercent)|>
  select(winpercent)|>
  head(5)
```

The least liked candies are Nik L Nip, Boston backed beans, Chiclets, supper bubble and jaw busters. 

```{r}
#Here we are using base r
head(candy[order(candy$winpercent),])
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(-winpercent)|>
  select(winpercent)|>
  head(5)
```
The top 5 are Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, Snickers

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r, fig.height=15, fig.width=10}
ggplot(candy)+
  aes(winpercent, rownames(candy))+
  geom_col()
```
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
Tip

```{r, fig.height=15, fig.width=10}
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy), winpercent))+
  geom_col()
```

## Adding Color 

```{r}
my_cols=rep("black",nrow(candy))
#Chocolate= Purple
my_cols[as.logical(candy$chocolate)]<- "purple"
#Bar= Yellow
my_cols[as.logical(candy$bar)]<- "green"
#Fruity= Red
my_cols[as.logical(candy$fruity)]<- "red"
```

```{r, fig.height=15, fig.width=10}
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy), winpercent))+
  geom_col(fill=my_cols)
```
> Q17. What is the worst ranked chocolate candy?

The Worse ranked chocolate candy is sixlets 

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is star bursts 

## Taking a look at pricepercent

Now lets compare the win percent versus the price of the candy (its value)

```{r}
ggplot(candy)+
  aes(winpercent, pricepercent, label =rownames(candy))+
  geom_point(color=my_cols)+
  geom_text()
```

Lets fix the label text over plotting using the an add-on called **ggrepel** and the `geom_text_repel()` function.

```{r}
library(ggrepel)
ggplot(candy)+
  aes(winpercent, pricepercent, label =rownames(candy))+
  geom_point(color=my_cols)+
  geom_text_repel(color=my_cols, max.overlaps=9)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

The best bang for your buck would be the reese's miniatures, you want to be low on the y-axis and high on the x axis. 

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
candy%>%
  arrange(-pricepercent)%>%
  select(pricepercent,winpercent)%>%
  head(n=5)
```
The top 5 most expensive Nik L Nip, Nestle smarties, Ring pop, Hershey's Krackel, Hershey Milk Chocolate. Of these 5 most expensive the most popular is Hershey's Krackel. 

## Exploring the correlation structure

We can calculate the pairwise correlation of all our columns.

```{r}
cij<- cor(candy)
library(corrplot)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Very negative (strongly anti-correlated) is fruit and chocolate indicating they don't coexist (normally).

> Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

Most likely PC1 will encompass whether the candy is chocolate or fruit based. This will have a lot of variance and split the data up nicely.

## Principal Component Analysis (PCA)

We should set the `scale=TRUE` due to the difference between the scaling in the `winpercent` and the other variables. if `scale=FALSE` this would dominate the PCA reuslts.

```{r}
pca<- prcomp(candy,scale=TRUE)
summary(pca)
```

"Score plot" of PC1 versus PC2, this will show how the different candies are related to one another on the separate PC axis. 

```{r}
ggplot(pca$x)+
  aes(PC1, PC2, label=row.names(pca$x))+
  geom_point(color=my_cols)+
  geom_text_repel(color=my_cols, max.overlaps = 7)
```

The second major results figure from PCA. This is called "loadings plot" --> how the original variables contribute to the PCA. 

```{r}
ggplot(pca$rotation)+
  aes(PC1,reorder(row.names(pca$rotation),PC1))+
  geom_col()
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

Picked up strongly is pluribus, fruity and hard. These all are highlighted in the correlation plot. They showed negative correlation in a lot of the variables within the correlation plot. 

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

There is a high correlation between a candy being chocolate and being popular. The win percents tend to be the highest amoungst groups with chocolates this can be seen in in the PCA of win percent and price percent, highlighted in green are chocolate candies with caramel, but these also tend to be the most expensive. If you are candy executive you would be able to charge the most and sell the most popular candy if it were chocolate.