---
title: "Class 5:Data Vis with ggplot2"
author: "Ryan Ellis (PID:A17673864)"
format: pdf
toc: TRUE

---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **ggplot2**. 

Let's compare how we make a simple figure with these two systems:

We can use the in-built `cars` dataset:

```{r}
head (cars)
plot(cars)
```

Before I can use ggplot 2 I need to install it on mu computer. To do this we can use the function `install.package("ggplot2")`

> **N.R** We never run `install.package()` in our quarto doc (we run it once only in the R console (R brain)), if we ran it in the doc it would reinstall the package everytime we render. 

once installed we need to load the package into our R brain: 

```{r}
library(ggplot2)
```


The main function in **ggplot 2** package is called `ggplot()`

```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers:

- the **data** (a data.frame of the stuff we want to plot)
- the **aes**thetics (how the data maps to the plot)
- the **geom** layer (how you want the map to be drawn, ie. points, lines, etc.)

```{r}
ggplot(cars)+
  aes(x=speed , y=dist )+
  geom_point()
```

## Add some custom features 

Let's add a trend line that shows a relationship between speed and distance

```{r}
ggplot(cars)+
  aes(x=speed , y=dist )+
  geom_point()+
  geom_line()
```

```{r}
ggplot(cars)+
  aes(x=speed , y=dist )+
  geom_point()+
  geom_smooth()+
  theme_bw()+
  labs(title= "Stopping Dist of old cars", 
       x="Speed(MPH)", 
       y="Dist(ft)")
```

Q. Can you make `geom_smooth()` produce a linear straight line of best fit for the data and turn off the "grey" error region.

---

## Gene Expression



```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A useful new function in this context is the `table()` function 

```{r}
table(genes$state)
```

My first Plot attempt

```{r}
ggplot(genes)+
  aes(Condition1, Condition2, col=State)+
  geom_point()+
  scale_color_manual (values=c("purple", "gray", "orange"))+
  theme_bw()+
  labs(x= "No Drug", y= "Drug", title= "Expression chnages upon GLP-1 inhibitor treatment")
```


## Going further

Here we read the famous gapmider data set:

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many entries (ie. rows) are there in the data set?

```{r}
length(table(gapminder$country))
```

```{r}
length(unique(gapminder$country))
```

Lets make our first plot of the entire dataset:

Plot of "gdpPercap" vs "lifeExp" colored by "continent"

```{r}
p <- ggplot(gapminder)+
  aes(gdpPercap,lifeExp, color=continent)+
  geom_point(alpha=0.3)
```

```{r}
p
```

```{r}
p+
  facet_wrap(~continent)
```

Make a plot for 1977 and 2007 only (not all years in the data set)

> Q. First use the **dplyr** package and the `filter()` function from that package to extract the rows from the year 2007

```{r,message=FALSE, warning=FALSE}
library(dplyr)
```

```{r}
g07<- filter(gapminder, year==2007)
g77<- filter(gapminder, year==1977)
g<- filter(gapminder, year==1977 | year==2007)
```

```{r}
ggplot(g07)+
  aes(gdpPercap,lifeExp, color=continent)+
  geom_point()
ggplot(g77)+
  aes(gdpPercap,lifeExp, color=continent)+
  geom_point()
ggplot(g)+
  aes(gdpPercap,lifeExp, color=continent)+
  geom_point()+
  facet_wrap(~year)

```

> Q.Make a histogram of lifeExp facetated by continent 
> Q.Make a histogram of lifeExp colored by continent (try using `fill=continent` or `col=continent`)

```{r,message=F}
ggplot(gapminder)+
  aes(lifeExp)+
  geom_histogram()
```

```{r,message=F}
ggplot(gapminder)+
  aes(lifeExp, col=continent)+
  geom_histogram()
```

```{r, message=F}
ggplot(gapminder)+
  aes(lifeExp)+
  geom_histogram()+
  facet_wrap(~continent)
```

```{r, message=F}
ggplot(gapminder)+
  aes(lifeExp, fill=continent)+
  geom_col()
```
